---
title: "M3S2 CW"
author: "Juliette Limozin"
date: " Due 09/03/2020"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(numbers)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stats)
library(tibble)
library(forecast)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

#Exploratory analysis of the data

```{r}
summary(dat)
summarise(dat, 'Rest pulse SD'= sd(rest_pulse), 'Stimulated pulse SD'= sd(stimulated_pulse), 'BMI SD'= sd(bmi))
ggplot(dat) +
  #geom_bar(aes(stimulated_pulse), colour = "red")+
  geom_bar(aes(bmi), colour = "blue") +
  labs(title = "Distribution of BMI")
```
##Correlation

```{r}
cat("Correlation between stimulated and rest pulses: \n",cor(dat$rest_pulse, dat$stimulated_pulse))
cat("\n")
cat("Correlation between stimulated pulse and BMI: \n",cor(dat$bmi, dat$stimulated_pulse))
```

#Clinicians' initial model

Includes measures and diagnostic plots
```{r}
fit0<-lm(dat$stimulated_pulse~dat$rest_pulse + dat$bmi)
summary(fit0)
plot(fit0)
anova(fit0)
cat("AIC: ",AIC(fit0))
cat("\n")
cat("Confidence interval of beta: \n")
print(confint.lm(fit0, level = 0.95))
ggplot(dat)+
  geom_jitter(aes(bmi, fit0$fitted.values, colour = "model")) +
  geom_jitter(aes(bmi, stimulated_pulse, colour = "data")) +
  labs( x = "BMI", y = "Stimulated Pulse",
   title ="Fitted vs Actual with Clinicians model") +
  scale_colour_manual("", values=c("model"="red", "data"="blue"), labels=mylabs)
ggplot(dat)+
  geom_jitter(aes(rest_pulse, fit0$fitted.values, colour = "model")) +
  geom_jitter(aes(rest_pulse, stimulated_pulse, colour = "data")) +
  labs( x = "Rest pulse", y = "Stimulated Pulse",
   title ="Fitted vs Actual with Clinicians model") +
  scale_colour_manual("", values=c("model"="red", "data"="blue"), labels=mylabs)
```
# Statistician's model

In-built version to sense check against the GLM I created manually: (includes diagnostic plots)
```{r in-built glm to sense-check}
y <- dat$stimulated_pulse-dat$rest_pulse
fit1<- glm(y~dat$bmi,family=Gamma(link="inverse"))
summary(fit1)
cat("AIC: ",AIC(fit1))
cat("\n")
cat("Confidence interval of beta: \n")
print(confint.lm(fit1, level = 0.95))
plot(fit1)
```

My version of the GLM:

```{r manual gamma GLM with inverse link function}

#Manually made GLM function
statglm <- function(d, tt = TRUE){ #inputs: d = dataset, tt is indicator for if we want a summary or not
  beta <- c(0.05,0) #initial estimate is 1/mean(Bmi), 0
  X <- cbind(1, d$bmi) #design matrix 
  y <- d$stimulated_pulse-d$rest_pulse #response output
  jj <- 0
  #Inverse of the link function (in this case it's the inverse function too)
  inv.link <-function(u){
    1/u
  }
  #Deviance function = 2*(l(y) - l(mu))
  D <- function(u){
    a = -1-log(y)
    b = y/u + log(u)
    2*sum(a+b)
  }
  oldD <- D(inv.link(as.numeric(X%*%beta))) #Current deviance
  while(jj ==0){
    eta = X%*%beta 
    mu <- inv.link(eta) #mu = 1/eta
    detadmu <- -1/(mu^2) #deta/dmu
    z <- eta +(y-mu)*detadmu #z
    w = mu^2 #weights
    lmod <- lm(z~d$bmi, weights=w) # regress z onto x with weights w
    beta <- as.vector(lmod$coeff) #new estimate of parameters beta
    newD <- D(inv.link(X%*%beta)) #new deviance
    control <- abs(newD-oldD)/(abs(newD)+0.1)
    if (isTRUE(control<1e-8)){
      jj<-1} #stop the algorithm when the control factor is small enough
    oldD <- newD
  }
  
  #Calculate statistics for diagnostics:
  J<- t(X)%*%diag(as.vector(w))%*%X
  invJ <-solve(J)
  beta.sd <- sqrt(as.vector(diag(invJ)))  #Standard errors
  t_value<- beta/beta.sd #t value
  p_value <- 2*pt(-abs(t_value), df=nrow(d)-ncol(d)) #p value
  errors <- as.vector(y - inv.link(X%*%beta)) #Residuals
  RSS <- t(errors)%*%errors #Residual sum of squares
  sst <- t(y-mean(y))%*%(y-mean(y)) #Total sum of squres
  R2 <- 1- RSS/sst #R-squared
  AR2 <- 1- (1-R2)*(nrow(d)-1)/(nrow(d)-ncol(X)) #Adjusted R-squared
  MSE <- RSS/(nrow(d)-ncol(X)) #MSE
  MSR <- (sst-RSS)/(ncol(X)-1) #MSR
  F_value <- MSR/MSE #F-statistic
  coeff <- data.frame(beta, beta.sd, t_value, p_value)
  dev <- c(sign(y[1]-mu[1])*sqrt(D(inv.link(X[1,]%*%beta))))
  for (i in 2:nrow(d)){
    dev <- c(dev,sign(y[i]-mu[i])*sqrt(D(inv.link(X[i,]%*%beta)))) #Deviance residuals
  }
  dev <- summary(dev) #Deviance residuals
  phi <- newD/(nrow(d)-2) #estimated dispersion parameters
  AIC <- -2*sum(dgamma(y, 1/phi, 1/(phi*mu), log = TRUE)) + 2*2 #AIC
  if (tt == TRUE){
    cat("Deviance residuals:\n")
    print(dev)
    cat("\n")
    cat("Coefficients:\n") 
    printCoefmat(coeff)
    cat("\n")
    cat("Residuals deviance: ",newD)
    cat("\n")
    cat("R-Squared: ", R2)
    cat(",\tAdjusted R-Squared: ", AR2,
	    "\nF-statistic:", F_value,
	    "with", nrow(d)-2-1, "DF")
    cat("\n")
    cat("AIC : ", AIC)
  }
  else{
  return(list("model" = inv.link(X%*%beta), "y" = y))}
}

statglm(dat, tt = TRUE)
```
Diagnostic plots

```{r,warning=FALSE}
ggplot(dat)+
  geom_jitter(aes(bmi, statglm(dat, tt = FALSE)$model, colour = "model")) +
  geom_jitter(aes(bmi, statglm(dat, tt = FALSE)$y, colour = "data")) +
  labs( x = "BMI", y = "Difference in Pulse",
   title ="Fitted vs Actual with Statisticians model") +
  scale_colour_manual("", values=c("model"="red", "data"="blue"), labels=mylabs)
```

#Binomial GLM

Modified version of the previous function

```{r,  warning=FALSE}
binomglm <- function(d, tt = TRUE){
  beta <- c(0.05,0)
  X <- cbind(1, d$bmi)
  y <- d$stimulated_pulse-d$rest_pulse
  jj <- 0
  inv.link <-function(u){
    nrow(d)*(1/(1+exp(-u)))
  }
  D <- function(u){
    a <- y*log(y/u)
    b = (nrow(d)-y)*log((nrow(d)-y)/(nrow(d)-u))
    a[y == 0] <- 0
    2*sum(a+b)
  }
  oldD <- D(inv.link(as.numeric(X%*%beta)))
  while(jj ==0){
    eta <- X%*%beta
    mu <- inv.link(eta)
    detadmu <- nrow(d)/(mu*(nrow(d)-mu))
    z <- eta +(y-mu)*detadmu
    w = mu*(nrow(d)-mu)/nrow(d)
    lmod <- lm(z~d$bmi, weights=w)
    beta <- as.vector(lmod$coeff)
    newD <- D(inv.link(X%*%beta))
    control <- abs(newD-oldD)/(abs(newD)+0.1)
    if (isTRUE(control<1e-8)){
      jj<-1}
    oldD <- newD
  }
  J<- t(X)%*%diag(as.vector(w))%*%X
  invJ <-solve(J)
  beta.sd <- sqrt(as.vector(diag(invJ)))
  t_value<- beta/beta.sd
  p_value <- 2*pt(-abs(t_value), df=nrow(d)-ncol(d))
  errors <- as.vector(y - inv.link(X%*%beta))
  RSS <- t(errors)%*%errors
  sst <- t(y-mean(y))%*%(y-mean(y))
  R2 <- 1- RSS/sst
  AR2 <- 1- (1-R2)*(nrow(d)-1)/(nrow(d)-2)
  MSE <- RSS/(nrow(d)-2)
  MSR <- (sst-RSS)/(2-1)
  F_value <- MSR/MSE
  coeff <- data.frame(beta, beta.sd, t_value, p_value)
  dev <- c(sign(y[1]-mu[1])*sqrt(D(inv.link(X[1,]%*%beta))))
  for (i in 2:nrow(d)){
    dev <- c(dev,sign(y[i]-mu[i])*sqrt(D(inv.link(X[i,]%*%beta))))
  }
  dev <- summary(dev)
  phi <- newD/(nrow(d)-2)
  AIC <- -2*sum(dbinom(y, nrow(d), as.vector(mu/nrow(d)), log = TRUE)) + 2*2
  if (tt == TRUE){
    cat("Deviance residuals:\n")
    print(dev)
    cat("\n")
    cat("Coefficients:\n") 
    printCoefmat(coeff)
    cat("\n")
    cat("Residuals deviance: ",newD)
    cat("\n")
    cat("R-Squared: ", R2)
    cat(",\tAdjusted R-Squared: ", AR2,
	    "\nF-statistic:", F_value,
	    "with", nrow(d)-2-1, "DF")
    cat("\n")
    cat("AIC : ", AIC)
  }
  else{
  return(list("model" = inv.link(X%*%beta), "y" = y))}
}

binomglm(dat, tt  = TRUE)

mylabs=list("Data", "Model") #labels for plots
#plot of model vs data
ggplot(dat)+
  geom_jitter(aes(bmi, binomglm(dat, tt = FALSE)$model, colour = "model")) +
  geom_jitter(aes(bmi, binomglm(dat, tt = FALSE)$y, colour = "data")) +
  labs( x = "BMI", y = "Difference in Pulse",
   title ="Fitted vs Actual with Statisticians model") +
  scale_colour_manual("", values=c("model"="red", "data"="blue"), labels=mylabs)


```


#Poisson GLM

Modified version of the previous function

```{r, warning=FALSE}
poissonglm <- function(d, tt = TRUE){
  beta <- c(0.05,0)
  X <- cbind(1, d$bmi)
  y <- d$stimulated_pulse-d$rest_pulse
  jj <- 0
  D <- function(u){
    2*sum(y*log(y/u)-y+u)
  }
  oldD <- D(as.numeric(X%*%beta))
  while(jj ==0){
    eta <- X%*%beta
    mu <- eta
    z <- eta +(y-mu)
    w = 1/mu
    lmod <- lm(z~d$bmi, weights=w)
    beta <- as.vector(lmod$coeff)
    newD <- D(X%*%beta)
    control <- abs(newD-oldD)/(abs(newD)+0.1)
    if (isTRUE(control<1e-8)){
      jj<-1}
    oldD <- newD
  }
  J<- t(X)%*%diag(as.vector(w))%*%X
  invJ <-solve(J)
  beta.sd <- sqrt(as.vector(diag(invJ)))
  t_value<- beta/beta.sd
  p_value <- 2*pt(-abs(t_value), df=nrow(d)-ncol(d))
  errors <- as.vector(y - (X%*%beta))
  RSS <- t(errors)%*%errors
  sst <- t(y-mean(y))%*%(y-mean(y))
  R2 <- 1- RSS/sst
  AR2 <- 1- (1-R2)*(nrow(d)-1)/(nrow(d)-2)
  MSE <- RSS/(nrow(d)-2)
  MSR <- (sst-RSS)/(2-1)
  F_value <- MSR/MSE
  coeff <- data.frame(beta, beta.sd, t_value, p_value)
  dev <- c(sign(y[1]-mu[1])*sqrt(D((X[1,]%*%beta))))
  for (i in 2:nrow(d)){
    dev <- c(dev,sign(y[i]-mu[i])*sqrt(D((X[i,]%*%beta))))
  }
  dev <- summary(dev)
  phi <- newD/(nrow(d)-2)
  AIC <- -2*sum(dpois(y, as.vector(mu), log = TRUE)) + 2*2
  if (tt == TRUE){
    cat("Deviance residuals:\n")
    print(dev)
    cat("\n")
    cat("Coefficients:\n") 
    printCoefmat(coeff)
    cat("\n")
    cat("Residuals deviance: ",newD)
    cat("\n")
    cat("R-Squared: ", R2)
    cat(",\tAdjusted R-Squared: ", AR2,
	    "\nF-statistic:", F_value,
	    "with", nrow(d)-2-1, "DF")
    cat("\n")
    cat("AIC : ", AIC)
  }
  else{
  return(list("model" = (X%*%beta), "y" = y, "Residuals" = errors))}
}

poissonglm(dat, tt  = TRUE)

mylabs=list("Data", "Model") #labels for plots
#plot of model vs data
ggplot(dat)+
  geom_jitter(aes(bmi, poissonglm(dat, tt = FALSE)$model, colour = "model")) +
  geom_jitter(aes(bmi, poissonglm(dat, tt = FALSE)$y, colour = "data")) +
  labs( x = "BMI", y = "Difference in Pulse",
   title ="Fitted vs Actual with Statisticians model") +
  scale_colour_manual("", values=c("model"="red", "data"="blue"), labels=mylabs)

```


#Gamma GLM with coffee added as covariate

```{r}
fit2 <- glm(y ~ dat$bmi + dat$coffee_reg, family = Gamma(link = "inverse"))
summary(fit2)
cat("Confidence interval of beta: \n")
print(confint.lm(fit2, level = 0.95))

#plot of model vs data
ggplot(dat)+
  geom_jitter(aes(bmi, fit2$fitted.values, colour = "model")) +
  geom_jitter(aes(bmi, stimulated_pulse - rest_pulse, colour = "data")) +
  labs( x = "BMI", y = "Difference in Pulse",
   title ="Fitted vs Actual with Statisticians model") +
  scale_colour_manual("", values=c("model"="red", "data"="blue"), labels=mylabs)

ggplot(dat)+
  geom_jitter(aes(coffee_reg, fit2$fitted.values, colour = "model")) +
  geom_jitter(aes(coffee_reg, stimulated_pulse - rest_pulse, colour = "data")) +
  labs( x = "BMI", y = "Difference in Pulse",
   title ="Fitted vs Actual with Statisticians model") +
  scale_colour_manual("", values=c("model"="red", "data"="blue"), labels=mylabs)

plot(fit2)
````


#I mverse gaussian GLM (not considered in the report)
```{r}
fit3 <- glm(y ~ dat$bmi + dat$coffee_reg, family = inverse.gaussian())
summary(fit3)
cat("Confidence interval of beta: \n")
print(confint.lm(fit3, level = 0.95))

ggplot(dat)+
  geom_jitter(aes(bmi, fit3$fitted.values, colour = "model")) +
  geom_jitter(aes(bmi, stimulated_pulse - rest_pulse, colour = "data")) +
  labs( x = "BMI", y = "Difference in Pulse",
   title ="Fitted vs Actual with Statisticians model") +
  scale_colour_manual("", values=c("model"="red", "data"="blue"), labels=mylabs)

ggplot(dat)+
  geom_jitter(aes(coffee_reg, fit3$fitted.values, colour = "model")) +
  geom_jitter(aes(coffee_reg, stimulated_pulse - rest_pulse, colour = "data")) +
  labs( x = "BMI", y = "Difference in Pulse",
   title ="Fitted vs Actual with Statisticians model") +
  scale_colour_manual("", values=c("model"="red", "data"="blue"), labels=mylabs)

plot(fit3)
````

#Gamma GLM with coffee added and log link function
```{r}
fit4 <- glm(y ~ dat$bmi + dat$coffee_reg, family = Gamma(link = "log"))
summary(fit4)
cat("Confidence interval of beta: \n")
print(confint.lm(fit4, level = 0.95))

ggplot(dat)+
  geom_jitter(aes(bmi, fit4$fitted.values, colour = "model")) +
  geom_jitter(aes(bmi, stimulated_pulse - rest_pulse, colour = "data")) +
  labs( x = "BMI", y = "Difference in Pulse",
   title ="Fitted vs Actual with Best model") +
  scale_colour_manual("", values=c("model"="red", "data"="blue"), labels=mylabs)

ggplot(dat)+
  geom_jitter(aes(coffee_reg, fit4$fitted.values, colour = "model")) +
  geom_jitter(aes(coffee_reg, stimulated_pulse - rest_pulse, colour = "data")) +
  labs( x = "Coffee factor", y = "Difference in Pulse",
   title ="Fitted vs Actual with Best model") +
  scale_colour_manual("", values=c("model"="red", "data"="blue"), labels=mylabs)

plot(fit4)
````